Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.3.2)

### 2016-01-21 MSstats_v3/MSstats.Rmd
2016.01.21-Meena, v3.3.2, separate folder for version

---
title: "Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.10)"
author: "Meena Choi, Tsung-Heng Tsai & Erik Verschueren"
date: "October 01, 2015"
output: pdf_document
bibliography: bibliography.bib
---

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstatsOverview.pdf)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/requiredInput.jpg)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/required_LCMS.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/required_DIA.png)

@@

```{r, eval=T}
setwd('/Users/Meena/Dropbox/MSstats_GitHub_document')
```

You can check your working directory by:

```{r, eval=T}
getwd()
```

@@

The first step in using the `MSstats` is to format the data as described in Section~2. We load here the prepared example dataset and confirm the data is stored in the correct format.

```{r, eval=T}
DDA2009.superhirn <- read.csv("RawData.DDA.csv")

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_QC_all.pdf)

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_QC_yeast.pdf)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_TMP_ProfilePlot_yeast.png)

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_TMP_ProfilePlot_wSummarization_yeast.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_TMP_random_ProfilePlot_wSummarization_yeast.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_VolcanoPlot_65.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_Heatmap_colorkey.pdf)

![](/Users/Meena/Dropbox/MSstats_GitHub_document/DDA2009_Heatmap.pdf)

@@

```{r, eval=T}
raw <- read.csv(file = "ControlMixerMSstatsInputfromskyline.csv")
head(raw)
```

@@

```{r, eval=T}
## 1. First, get protein ID information
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

```{r, eval=T}
## 2. Read in annotation including condition and biological replicates: annotation.csv
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

@@

```{r, eval=T, echo=FALSE}
load("DDA2014.evidence.RData")
infile <- DDA2014.evidence
```

```{r, eval=T, warning=F}
## 4. Reformat for MSstats required input
##    check options for converting format
?MaxQtoMSstatsFormat
```

```{r, eval=T}
msstats.raw <- MaxQtoMSstatsFormat(evidence=infile, annotation=annot, proteinGroups=proteinGroups)

## now 'msstats.raw' is ready for MSstats
head(msstats.raw)
```

>>>>

---
title: "Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.10)"
author: "Meena Choi, Tsung-Heng Tsai & Erik Verschueren"
date: "January 21, 2016"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.3.2)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
bibliography: bibliography.bib
---

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/MSstatsOverview.pdf)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/requiredInput.jpg)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/required_LCMS.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/required_DIA.png)

@@

```{r, eval=F}
setwd('/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3')
```

@@

The first step in using the `MSstats` is to format the data as described in Section 2. We load here the prepared example dataset and confirm the data is stored in the correct format.

```{r, eval=T}
DDA2009.superhirn <- DDARawData

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_QC_all.pdf)

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_QC_yeast.pdf)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_TMP_ProfilePlot_yeast.png)

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_TMP_ProfilePlot_wSummarization_yeast.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_TMP_random_ProfilePlot_wSummarization_yeast.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_VolcanoPlot_65.png)

@@

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_Heatmap_colorkey.pdf)

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_Heatmap.pdf)

@@

```{r, eval=T}
raw<-DDARawData.Skyline
```

@@

```{r, eval=F}
## 1. First, get protein ID information
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

```{r, eval=F}
## 2. Read in annotation including condition and biological replicates: annotation.csv
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

@@

```{r, eval=F, echo=FALSE}
load("DDA2014.evidence.RData")
infile <- DDA2014.evidence
```

```{r, eval=F, warning=F}
## 4. Reformat for MSstats required input
##    check options for converting format
?MaxQtoMSstatsFormat
```

```{r, eval=F}
msstats.raw <- MaxQtoMSstatsFormat(evidence=infile, annotation=annot, proteinGroups=proteinGroups)

## now 'msstats.raw' is ready for MSstats
```

### 2016-01-21 MSstats_v3/MSstats_manual_RMarkdown.Rmd
2016.01.21-Meena, v3.3.2, separate folder for version

---
title: "Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.10)"
author: "Meena Choi, Tsung-Heng Tsai & Erik Verschueren"
date: "January 21, 2016"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.3.2)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
bibliography: bibliography.bib
---

@@

```{r, eval=F}
install.packages(pkgs = 'MSstats_3.0.9.tar.gz', repos = NULL, type = 'source')
```

@@

```{r, eval=F}
setwd('/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3')
```

@@

The first step in using the `MSstats` is to format the data as described in Section 2. We load here the prepared example dataset and confirm the data is stored in the correct format.

```{r, eval=T}
DDA2009.superhirn <- DDARawData

@@

A typical label-free DDA dataset may have many missing values and noisy features with outliers. `MSstats` supports several ways to deal with this. The default option for summarization is `TMP` (Tukey's median polish, robust parameter estimation method with median across rows and columns) after imputation by AFT (accelerated failure time model, `MBimpute=TRUE`) based on censored intensity for `NA` (`censoredInt="NA"`) with a cutoff as the minimum value for a feature and a run (`cutoffCensored="minFeatureNRun"`). Here the `NA` values represent censored measurements, which are below the cutoff and undetectable. This process handles missing values through imputation and reduces the influence of the outliers using the `TMP` estimation. Note, however, that those runs with no measurements at all will be removed and not be used for any calculation.

```{r, eval=T}
# default option
DDA2009.TMP <- dataProcess(raw = DDA2009.superhirn,  fillIncompleteRows = TRUE,
                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt = "NA", cutoffCensored = "minFeatureNRun",
                           MBimpute = TRUE)
```

If `MBimpute=FALSE`, imputation will not be performed and censored intensities (here `NA`'s) will be replaced with a cutoff value specified in `cutoffCensored`. The default option for `cutoffCensored` is `minFeatureNRun`, taking the minimum value for the corresponding feature and run. With this option, those runs with substantial missing measurements will be biased by the cutoff value. In such case, you may remove the runs that have more than 50% missing values from the analysis with the option `remove50missing=TRUE`.

@@

```{r, eval=F}
groupComparisonPlots(data = DDA2009.comparisons$ComparisonResult, type = 'VolcanoPlot')
```

'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison defined in `contrast.matrix`. Below is one of volcano plots, for comparison 'C6-C5'. Please check `?groupComparisonPlots` for detail, such as labelling protein names, size of dots, font sizes, etc.

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_VolcanoPlot_65.png)

@@

```{r, eval=T}
raw<-DDARawData.Skyline
```

@@

raw$pepprecursor<-paste(raw$PeptideSequence, raw$PrecursorCharge, sep="_")

data_w = dcast( Run ~ pepprecursor, data=raw, value.var='Intensity', fun.aggregate=sum, fill=NULL)

newdata = melt(data_w, id.vars=c('Run'))
colnames(newdata)[colnames(newdata) %in% c("variable","value")]<-c('pepprecursor','Intensity')

uniinfo<-unique(raw[,c("ProteinName","PeptideSequence","PrecursorCharge","pepprecursor")])		
newraw<-merge(newdata,uniinfo, by="pepprecursor")

uniinfo<-unique(raw[,c("Run","BioReplicate","Condition")])		
newraw<-merge(newraw,uniinfo, by="Run")

newraw$BioReplicate<-1 # it should be change based on your experiment.
newraw$FragmentIon<-"sum"
newraw$ProductCharge<-NA
newraw$IsotopeLabelType<-"L"

raw<-newraw
# now 'raw' is ready to use MSstats
```

@@

```{r, eval=F}
## 1. First, get protein ID information
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

```{r, eval=F}
## 2. Read in annotation including condition and biological replicates: annotation.csv
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

@@

```{r, eval=F, echo=FALSE}
load("DDA2014.evidence.RData")
infile <- DDA2014.evidence
```

```{r, eval=F, warning=F}
## 4. Reformat for MSstats required input
##    check options for converting format
?MaxQtoMSstatsFormat
```

```{r, eval=F}
msstats.raw <- MaxQtoMSstatsFormat(evidence=infile, annotation=annot, proteinGroups=proteinGroups)

## now 'msstats.raw' is ready for MSstats
```

>>>>

---
title: "Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.3.2)"
author: "Meena Choi, Tsung-Heng Tsai & Erik Verschueren"
date: "January 21, 2016"
output: pdf_document
bibliography: bibliography.bib
---

@@

```{r, eval=F}
install.packages(pkgs = 'MSstats_3.3.2.tar.gz', repos = NULL, type = 'source')
```

@@

```{r, eval=T}
setwd('/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3')
```

You can check your working directory by:

```{r, eval=T}
getwd()
```

@@

The first step in using the `MSstats` is to format the data as described in Section~2. We load here the prepared example dataset and confirm the data is stored in the correct format.

```{r, eval=T}
DDA2009.superhirn <- read.csv("RawData.DDA.csv")

@@

A typical label-free DDA dataset may have many missing values and noisy features with outliers. `MSstats` supports several ways to deal with this. The default option for summarization is `TMP` (Tukey's median polish, robust parameter estimation method with median across rows and columns) after imputation by AFT (accelerated failure time model, `MBimpute=TRUE`) based on censored intensity for `NA` (`censoredInt="NA"`) with a cutoff as the minimum value for a feature (`cutoffCensored="minFeature"`). Here the `NA` values represent censored measurements, which are below the cutoff and undetectable. This process handles missing values through imputation and reduces the influence of the outliers using the `TMP` estimation. Note, however, that those runs with no measurements at all will be removed and not be used for any calculation.

```{r, eval=T}
# default option
DDA2009.TMP <- dataProcess(raw = DDA2009.superhirn,  fillIncompleteRows = TRUE,
                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt = "NA", cutoffCensored = "minFeature",
                           MBimpute = TRUE)
```

If `MBimpute=FALSE`, imputation will not be performed and censored intensities (here `NA`'s) will be replaced with a cutoff value specified in `cutoffCensored`. The default option for `cutoffCensored` is `minFeature`, taking the minimum value for the corresponding feature. With this option, those runs with substantial missing measurements will be biased by the cutoff value. In such case, you may remove the runs that have more than 50% missing values from the analysis with the option `remove50missing=TRUE`.

***NOTE***
In case that there are completely missing measurements in a run for a protein, any imputation will not be performed. In addition, the condition, which has no measurement at all in a protein, will be not imputed.

@@

```{r, eval=F}
groupComparisonPlots(data = DDA2009.comparisons$ComparisonResult, type = 'VolcanoPlot', 
                     width=8, height=8)
```

'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison defined in `contrast.matrix`. Please check `?groupComparisonPlots` for detail, such as labelling protein names, size of dots, font sizes, etc. Below is one of volcano plots, for comparison 'C3-C2' including protein name labelling. Protein name will be shown for significant proteins and higher fold change if you set up  `FCcutoff`, without overlapping protein names each other.

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstats_v3/DDA2009_VolcanoPlot_23.png)

@@

```{r, eval=T}
raw <- read.csv(file = "ControlMixerMSstatsInputfromskyline.csv")
head(raw)
```

@@

raw$pepprecursor <- paste(raw$PeptideSequence, raw$PrecursorCharge, sep="_")

data_w <- dcast( Run ~ pepprecursor, data=raw, value.var='Intensity', fun.aggregate=sum, fill=NULL)

newdata <- melt(data_w, id.vars=c('Run'))
colnames(newdata)[colnames(newdata) %in% c("variable", "value")] <- c('pepprecursor', 'Intensity')

uniinfo <- unique(raw[, c("ProteinName", "PeptideSequence", "PrecursorCharge", "pepprecursor")])		
newraw <- merge(newdata, uniinfo, by="pepprecursor")

uniinfo <- unique(raw[, c("Run", "BioReplicate", "Condition")])		
newraw <- merge(newraw, uniinfo, by="Run")

newraw$BioReplicate <- 1 # it should be change based on your experiment.
newraw$FragmentIon <- "sum"
newraw$ProductCharge <- NA
newraw$IsotopeLabelType <- "L"

raw <- newraw
# now 'raw' is ready to use MSstats
```

@@

```{r, eval=T}
## 1. First, get protein ID information
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

```{r, eval=T}
## 2. Read in annotation including condition and biological replicates: annotation.csv
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

@@

```{r, eval=T, echo=FALSE}
load("DDA2014.evidence.RData")
infile <- DDA2014.evidence
```

```{r, eval=T, warning=F}
## 4. Reformat for MSstats required input
##    check options for converting format
?MaxQtoMSstatsFormat
```

```{r, eval=T}
msstats.raw <- MaxQtoMSstatsFormat(evidence=infile, annotation=annot, proteinGroups=proteinGroups)

## now 'msstats.raw' is ready for MSstats
head(msstats.raw)
```
